\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-26 paper.tex
Edits from Yan, Jere

\section{Event ARGs}
\label{sec-eARG}
% Trying to be totally, boringly, abundantly, blatantly clear here what we're
% talking about.
In this section we define the classical view of an ARG data structure,
and illustrate its limitations. We are
interested in the details of how ARGs are described mathematically,
and as a consequence, how they are represented in a practical sense
as the output of inference programs.
Where such concrete details of an ARG data structure
(the encoding) are
% TODO more?
mathematically defined~\citep[e.g.][]{
rasmussen2014genome,gusfield2014recombinatorics,ignatieva2021kwarg}
they follow the approach described by Griffiths and colleagues
(but see~\citet{parida2011minimal} and \citet{zhang2023biobank}
for notable exceptions), and the majority of
% Note the "an" here to avoid cases like ARGweaver
ARG inference methods use it as an output format
\citep[(TODO exhaustive list)][]{rasmussen2014genome,ignatieva2021kwarg}.
In this Griffiths encoding, we have two different types of node in the graph,
representing the common ancestor and recombination events
in the history of a sample.
At common ancestor nodes, the inbound lineages merge into a
single ancestral lineage with one parent, and at recombination
nodes a single lineage is split into two independent
ancestral lineages. Recombination nodes are annotated with
the corresponding crossover breakpoints. These breakpoints
are used to construct the local trees.
This is done by tracing rootwards through the graph from the samples,
making decisions about which outbound edge to follow through
recombination nodes based on the breakpoint
position~\citep{griffiths1996ancestral}.
Because it is focused on recording events and their properties,
we will refer to this Griffiths encoding as the ``event ARG'', or eARG
encoding. Fig.~\ref{fig-event-arg} shows an example of a classical
eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c})
and a single recombination event (node \noderef{d}) with a breakpoint
at position $x$.
We assume here that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.

@@

\end{tikzpicture}
\caption{\label{fig-event-arg}
A classical event ARG. (A) Standard graph depiction with
breakpoint $x$ associated with the recombination node \noderef{d}.
Nodes \noderef{e}, \noderef{f} and \noderef{g} are common ancestor events.
(B) Corresponding local trees to the left and right of breakpoint $x$
(note these are shown in the conventional form in which only coalescences
within the local tree are included; see Section~\ref{sec-ARG-and-local-trees}
% TODO: check above if we want to refer to sec-ARG-and-local-trees or some other!
for a discussion of this important point).
}
\end{figure}

% eARGs are representationally for limited interchange
While the Griffiths approach of annotating recombination nodes with a
breakpoint in an eARG is a concise and elegant way of describing realisations
of the coalescent, it has some significant issues when applied to the
complexities of present-day datasets. The eARG encoding explicitly models only
two different types of event and thus anything that is not a single crossover
recombination or common ancestor event, must be incorporated
either in a roundabout way using these
events, or by adding new types of event to the encoding. For example, gene
conversion could be accommodated either by stipulating a third type of event
(annotated by two breakpoints and corresponding traversal conventions for
recovering the local trees) or by two recombination nodes joined by a
zero-length edge. From the perspective of practical interchange of data between
inference methods and downstream applications, both
workarounds are problematic, and the gARG encoding described in the previous
section offers a much simpler solution.
Modern datasets contain many more complexities which the eARG encoding
is ill-suited to represent (see Discussion).

@@

\section{Ancestral material and sample resolution}
\label{sec-ancestral-material}
Ancestral material~\citep{wiuf1999ancestry,wiuf1999recombination}
is a key concept in understanding the overall inheritance structure
of an ARG. It denotes the genomic intervals ancestral to a set of samples
on the edges of an ARG.
In Fig.~\ref{fig-arg-in-pedigree}, for example, we have
four sample genomes, \noderef{a}--\noderef{d}. As we
trace their genetic ancestry into the previous generation
(\noderef{e}--\noderef{h}), we can think of the ancestral
material of the samples propagating through the graph
backwards in time. In the interval $[2, 7)$, there is a
local coalescence where nodes \noderef{a} and \noderef{c}
find a common ancestor in \noderef{f}. Thus, in this interval,
the total number of genome segments that are ancestral to the
sample is reduced from four to three. Note that in the same
interval, node \noderef{e} is not ancestral to any
of the samples: it carries non ancestral material.
This process of local coalescence continues through the
graph, until all samples reach their most recent common
ancestor in node \noderef{n}.

The process of tracking local coalescences and updating
segments of ancestral material is a core element of
Hudson's seminal simulation
algorithm~\citep{hudson1983testing,kelleher2016efficient},
and the key distinguishing feature between the
``Big'' and ``Little'' ARG stochastic processes
(see Appendix~\ref{sec-big-and-little-arg}).
The ability to \emph{store} the resolved ancestral material
for a sample
in an ARG is also a key distinction between the eARG and gARG
encodings. Because an eARG stores only the graph topology and
recombination breakpoints, there is no way to locally
ascertain the ancestral material without traversing the graph
rootwards from the samples,
resolving the effects of recombination and common ancestor events.

Efficiently propagating and resolving ancestral material for
a sample through a pre-existing graph is a well-studied problem,
and central to recent advances in individual-based forwards-time
simulations~\citep{kelleher2018efficient,haller2018tree}.
In contrast to the usual ``retrospective'' view of ARGs
discussed so far, these methods record an ARG forwards in
time in a ``prospective'' manner. Genetic inheritance relationships
are recorded exhaustively, generation-by-generation,
leading to a rapid build-up of information, much of which
will not be relevant to the genetic ancestry of the current population.
This redundancy is periodically removed using the ``simplify''
algorithm~\citep{kelleher2018efficient}, which propagates and
resolves ancestral material.
Simplification allows efficient subsetting of ARGs (removing both samples and
certain levels of detail about recombination, as discussed in
Section~\ref{sec-ARG-simplification}), and is also the key enabling factor when
incorporating propective ARGs into
forward-time simulation.

@@

This distinction between ARG and local tree arity is mainly
of interest when we consider unary nodes, those that have
a single child.
Returning to the example in Fig~\ref{fig-arg-in-pedigree}, nodes
\noderef{g} and \noderef{h}
are ARG-unary (Fig~\ref{fig-arg-in-pedigree}B), and are consequently
also unary in the local trees (Fig~\ref{fig-arg-in-pedigree}C).
On the other hand, node \noderef{f} has two children
in the graph, but is binary only
in the local tree covering the interval $[2, 7)$,
representing the coalescence of samples \noderef{a} and \noderef{c}
in this genome region. Over the interval $[0, 2)$ no coalescence occurs,
but we still record the fact that genome \noderef{c} inherits from \noderef{f}
in the local tree. Thus, \noderef{f} has a single child in this
interval and it is unary in this local tree: in other words is is a \emph{locally unary} node.
% TODO note that node e is CA but not coalescent, and refer to where we discuss this.
In another example, \noderef{e} is binary in the graph, but is
locally unary in all local trees in which it is present.
This is because \noderef{a} and \noderef{c} are both recombinant
offspring of \noderef{e} and \noderef{f}, and,
in this example,
\noderef{a} inherits from
\noderef{e} over the interval $[0, 2)$,
and \noderef{c} inherits from \noderef{e} over $[7, 10)$.

@@

% Here we move on to coalescence
As for inferred coalescences, the shading (``coalescent span'')
reveals that many non-recombinant (green) nodes are
locally unary (see section \ref{sec-locally-unary-edges}). In other words,
in some regions of the genome these nodes are present in the ancestry without
corresponding to coalescence of ancestral material. As argued previously, this can provide useful
information about shared ancestry. Note however, that such unary segments are not inferred
in the \relate\ ARG: other than the samples it has entirely solid nodes.
Another clear difference involves coalescent nodes with more than two children (polytomies);
in Appendix \ref{sec-cell-lineages-and-args} we argue that this reflects
uncertainty in the order of coalescent events, in an equivalent
manner to the imprecision associated with recombination nodes having more
than two parents. Only \tsinfer\ creates such polytomies, the clearest of which
involves the nodes \noderef{Fr-F}, \noderef{Wa-F}, and \noderef{Af-F}. These
happen to have identical sequences, but because \kwarg, \argweaver, and \relate\ require
bifurcating trees, they are forced to pick an arbitrary order of branching
(hence Fig.~\ref{fig-inferred-args}A,B disagrees with Fig.~\ref{fig-inferred-args}D
in this order).
More generally, the number of nodes in Fig.~\ref{fig-inferred-args}C
that have more than two children hints at considerable uncertainty in
the order of branching elsewhere in the ARG.

@@

\section{The Big and Little ARG}
\label{sec-big-and-little-arg}
Here we review two important stochastic processes that construct ARGs:
the ``Big" ARG process of \cite{griffiths1997ancestral}, and the ``Little" ARG process of
 \cite{hudson1983properties}. The Big ARG process is mathematically simpler
 but is computationally intractable due to generating a vast number of ancestors
 which contribute no genetic material to the initial sample.
The Little ARG process avoids non-genetic ancestors at the cost of more complex
dynamics and state space. We also demonstrate that evaluating the sampling probability
of either process---a key quantity in many statistical approaches---requires that the
gARG (or eARG) data structure be interpreted in a model-specific way.

@@

	% Gray dashed lines to visually separate holding times
	\draw[color=gray, dashed] (7.5, 1.2) -- (9.1, 1.2);
	\draw[color=gray, dashed] (7.5, 1.7) -- (9.1, 1.7);
	\draw[color=gray, dashed] (7.5, 2.4) -- (9.1, 2.4);
	\draw[color=gray, dashed] (7.5, 3.1) -- (9.1, 3.1);
	\draw[color=gray, dashed] (7.5, 3.8) -- (9.1, 3.8);
	\draw[color=gray, dashed] (7.5, 4.5) -- (9.1, 4.5);
	\draw[color=gray, dashed] (7.5, 5.4) -- (9.1, 5.4);
	\draw[color=gray, dashed] (7.5, 6) -- (9.1, 6);
	\draw[color=gray, dashed] (7.5, 6.6) -- (9.1, 6.6);
	\draw[color=gray, dashed] (7.5, 7.2) -- (9.1, 7.2);
\end{tikzpicture}
\end{tabular}
}
\caption{(A)
A realisation of the graph traversed by Hudson's algorithm started from a
sample of three chromosomes with $m$ discrete sites each at time $t = 0$, and
propagated until time $T$. The MRCA on the genetic interval $[v, w)$ is reached
at event \noderef{b}, while that on $[0, v)$ is reached at event \noderef{c}.
The non-ancestral segment $[v, w)$ above
\noderef{a} contributes to the rate of effective recombinations because it
is trapped between ancestral segments. The two columns titled CA and RE
are the respective rates of mergers and recombinations when
the recombination rate is $\rho$.
(B) A corresponding realisation of a Big ARG, which augments Hudson's algorithm
by tracking nonancestral lineages. The result is a simpler state space and
dynamics, at the cost of extra nodes and edges, highlighted in red, which do
not affect the local tree at any site.
Recombination positions are labelled alphabetically in time, and their ordering
along the genome is $y < v < x < w$, of which the first only appears in panel B.
There are two separate recombination events at link $w$.
}
\label{hudson_vs_bigARG}
\end{figure}

Figure \ref{hudson_vs_bigARG} illustrates the more complex state space
of the Little ARG process, as well as the extra events which occur in the Big ARG process.
Moreover, it depicts the rates of common ancestors and recombination events in each
interval of time of the realisations.
In order to evaluate these rates, and hence the sampling probability
\cite[see e.g.][Equation (3)]{mahmoudi2022bayesian},
it is necessary to know the number of lineages and number of extant links
available for recombination in each time interval.
Some representations may not provide this information.
For example, in the gARG encoding depicted in Figure \ref{fig-ancestry-resolution}C
it is clear that a recombination takes place between nodes \noderef{b} and
\noderef{d} as well as \noderef{e}, but the exact time of the event is ambiguous,
and thus so is the number of lineages during the time interval. In fact, this
information cannot be recovered from the gARG encoding used in Figure \ref{fig-ancestry-resolution}C,
and alternative conventions must be introduced to resolve such ambiguities.
For the likelihood-based inference algorithms for the coalescent with recombination,
it is sufficient to create \emph{two} gARG nodes at the time of the recombination event:
see the appendix of \cite{baumdicker2021efficient} for details of evaluating
coalescent with recombination likelihoods using this convention.
This is also the interpretation depicted in
Figure \ref{hudson_vs_bigARG}, but it means that the two edges above node \noderef{b}
in Figure \ref{fig-ancestry-resolution} should correspond to only one lineage,
along which all $m-1$ links are available for recombination.
The lineage then splits into two at the time of nodes \noderef{d} and \noderef{e}.
Nodes \noderef{k}, \noderef{l}, and \noderef{m} in Figure \ref{fig-ancestry-resolution}
demonstrate that the same issue can affect the number of links available for recombination:
without an external convention, the exact time at which the trapped ancestral material on
node \noderef{k} ceases to be available for an effective recombination in the Little ARG process.

>>>>

\section{Event ARGs}
\label{sec-eARG}
% Trying to be totally, boringly, abundantly, blatantly clear here what we're
% talking about.
In this section we define the classical view of an ARG data structure,
and illustrate its limitations. We are
interested in the details of how ARGs are described mathematically,
and as a consequence, how they are represented in a practical sense
as the output of inference programs.
Where such concrete details of an ARG data structure
(the encoding) are
% TODO more?
mathematically defined~\citep[e.g.][]{
rasmussen2014genome,gusfield2014recombinatorics,ignatieva2021kwarg}
they follow the approach described by Griffiths and colleagues
(but see~\citet{parida2011minimal} and \citet{zhang2023biobank}
for notable exceptions), and the majority of
% Note the "an" here to avoid cases like ARGweaver
ARG inference methods use it as an output format
\citep[(TODO exhaustive list)][]{rasmussen2014genome,ignatieva2021kwarg}.
In this Griffiths encoding, as well as sample nodes,
we have two additional types of node in the graph,
representing the common ancestor and recombination events
in the history of a sample.
At common ancestor nodes, the inbound lineages merge into a
single ancestral lineage with one parent, and at recombination
nodes a single lineage is split into two independent
ancestral lineages. Recombination nodes are annotated with
the corresponding crossover breakpoints. These breakpoints
are used to construct the local trees.
This is done by tracing rootwards through the graph from the samples,
making decisions about which outbound edge to follow through
recombination nodes based on the breakpoint
position~\citep{griffiths1996ancestral}.
Because it is focused on recording events and their properties,
we will refer to this Griffiths encoding as the ``event ARG'' or eARG
encoding. Fig.~\ref{fig-event-arg} shows an example of a classical
eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c})
and a single recombination event (node \noderef{d}) with a breakpoint
at position $x$.
We assume here that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.

@@

\end{tikzpicture}
\caption{\label{fig-event-arg}
A classical event ARG (eARG). (A) Standard graph depiction with
breakpoint $x$ associated with the recombination node \noderef{d}.
Nodes \noderef{e}, \noderef{f} and \noderef{g} are common ancestor events.
(B) Corresponding local trees to the left and right of breakpoint $x$
(note these are shown in the conventional form in which only coalescences
within the local tree are included; see Section~\ref{sec-ARG-and-local-trees}
% TODO: check above if we want to refer to sec-ARG-and-local-trees or some other!
for a discussion of this important point).
}
\end{figure}

% eARGs are representationally for limited interchange
While the Griffiths approach of annotating recombination nodes with a
breakpoint in an eARG is a concise and elegant way of describing realisations
of the coalescent with recombination, it has some significant issues when applied to the
complexities of present-day datasets. The eARG encoding explicitly models only
two different types of event and thus anything that is not a single crossover
recombination or common ancestor event, must be incorporated
either in a roundabout way using these
events, or by adding new types of event to the encoding. For example, gene
conversion could be accommodated either by stipulating a third type of event
(annotated by two breakpoints and corresponding traversal conventions for
recovering the local trees) or by two recombination nodes joined by a
zero-length edge. From the perspective of practical interchange of data between
inference methods and downstream applications, both
workarounds are problematic, and the gARG encoding described in the previous
section offers a much simpler solution.
Modern datasets contain many more complexities which the eARG encoding
is ill-suited to represent (see Discussion).

@@

\section{Ancestral material and sample resolution}
\label{sec-ancestral-material}
Ancestral material~\citep{wiuf1999ancestry,wiuf1999recombination}
is a key concept in understanding the overall inheritance structure
of an ARG. It denotes the genomic intervals ancestral to a set of samples
on the edges of an ARG.
For example, in Fig.~\ref{fig-arg-in-pedigree} we have
four sample genomes, \noderef{a}--\noderef{d}. As we
trace their genetic ancestry into the previous generation
(\noderef{e}--\noderef{h}), we can think of their ancestral
material propagating through the graph
backwards in time. In the region $[2, 7)$, there is a
local coalescence where nodes \noderef{a} and \noderef{c}
find a common ancestor in \noderef{f}. Thus, in this region,
the total number of genome segments that are ancestral to the
sample is reduced from four to three. Fig.~\ref{fig-arg-in-pedigree}A
illustrates this by (shaded) ancestral material being present
in only three nodes (\noderef{f}, \noderef{g}, and \noderef{h}) in this region,
while node \noderef{e} is blank
as it carries \emph{non ancestral} material.
This process of local coalescence continues through the
graph, until all samples reach their most recent common
ancestor in node \noderef{n}.

The process of tracking local coalescences and updating
segments of ancestral material is a core element of
Hudson's seminal simulation
algorithm~\citep{hudson1983testing,kelleher2016efficient},
and the key distinguishing feature between the
``Big'' and ``Little'' ARG stochastic processes
(see Appendix~\ref{sec-big-and-little-arg}).
The ability to \emph{store} resolved ancestral material
is also a key distinction between the eARG and gARG
encodings. Because an eARG stores only the graph topology and
recombination breakpoints, there is no way to locally
ascertain ancestral material without traversing the graph
rootwards from the sample nodes,
resolving the effects of recombination and common ancestor events.

Efficiently propagating and resolving ancestral material for
a sample through a pre-existing graph is a well-studied problem,
and central to recent advances in individual-based forward-time
simulations~\citep{kelleher2018efficient,haller2018tree}.
In contrast to the usual ``retrospective'' view of ARGs
discussed so far, these methods record an ARG forwards in
time in a ``prospective'' manner. Genetic inheritance relationships
are recorded exhaustively, generation-by-generation,
leading to a rapid build-up of information, much of which
will not be relevant to the genetic ancestry of the current population.
This redundancy is periodically removed using the ``simplify''
algorithm~\citep{kelleher2018efficient}, which propagates and
resolves ancestral material.
Simplification allows efficient subsetting of ARGs (removing both samples and
certain levels of detail about recombination, as discussed in
Section~\ref{sec-ARG-simplification}), and is also the key enabling factor when
incorporating prospective ARGs into
forward-time simulation.

@@

This distinction between ARG and local tree arity is mainly
of interest when we consider those that have
a single child: that are \emph{unary}.
Returning to the example in Fig~\ref{fig-arg-in-pedigree}, nodes
\noderef{g} and \noderef{h}
are ARG-unary (Fig~\ref{fig-arg-in-pedigree}B), and are consequently
also unary in the local trees (Fig~\ref{fig-arg-in-pedigree}C).
On the other hand, node \noderef{f} has two children
in the graph, but is binary only
in the local tree covering the interval $[2, 7)$,
representing the coalescence of samples \noderef{a} and \noderef{c}
in this genome region. Over the interval $[0, 2)$ no coalescence occurs,
but we still record the fact that genome \noderef{c} inherits from \noderef{f}
in the local tree. Thus, \noderef{f} has a single child in this
interval and it is unary in this local tree: in other words it is a \emph{locally unary} node.
% TODO note that node e is CA but not coalescent, and refer to where we discuss this.
In another example, \noderef{e} is binary in the graph, but is
locally unary in all trees in which it is present.
This is because \noderef{a} and \noderef{c} are both recombinant
offspring of \noderef{e} and \noderef{f}, and,
in this example,
\noderef{a} inherits from
\noderef{e} over the interval $[0, 2)$,
and \noderef{c} inherits from \noderef{e} over $[7, 10)$.

@@

% Here we move on to coalescence
As for inferred coalescences, the shading (``coalescent span'')
reveals that many non-recombinant (green) nodes are
locally unary (see Section \ref{sec-locally-unary-edges}). In other words,
in some regions of the genome these nodes are present in the ancestry without
corresponding to coalescence of ancestral material. As argued previously, this can provide useful
information about shared ancestry. Note however, that such unary segments are not inferred
in the \relate\ ARG: other than the samples it has entirely solid nodes.
Another clear difference involves coalescent nodes with more than two children (polytomies);
in Appendix \ref{sec-cell-lineages-and-args} we argue that this reflects
uncertainty in the order of coalescent events, in an equivalent
manner to the imprecision associated with recombination nodes having more
than two parents. Only \tsinfer\ creates such polytomies, the clearest of which
involves the nodes \noderef{Fr-F}, \noderef{Wa-F}, and \noderef{Af-F}. These
happen to have identical sequences, but because \kwarg, \argweaver, and \relate\ require
bifurcating trees, they are forced to pick an arbitrary order of branching
(hence Fig.~\ref{fig-inferred-args}A,B disagrees with Fig.~\ref{fig-inferred-args}D
in this order).
More generally, the number of nodes in Fig.~\ref{fig-inferred-args}C
that have more than two children hints at considerable uncertainty in
the order of branching elsewhere in the ARG.

@@

\section{The Big and Little ARG}
\label{sec-big-and-little-arg}
Here we review two important stochastic processes that construct ARGs:
the ``Big" ARG process of \cite{griffiths1997ancestral}, and the ``Little" ARG process of
 \cite{hudson1983properties}. The Big ARG process is mathematically simpler
 but is computationally intractable due to generating a vast number of ancestors
 which contribute no genetic material to the initial sample.
The Little ARG process avoids non-genetic ancestors at the cost of more complex
dynamics and state space. We also demonstrate that applications relying on the grouping of inheritance pathways into ancestral lineages, such as likelihood-based inference under the coalescent, requires that the
gARG (or eARG) data structure be interpreted in a model-specific way.

@@

	% Gray dashed lines to visually separate holding times
	\draw[color=gray, dashed] (7.5, 1.2) -- (9.1, 1.2);
	\draw[color=gray, dashed] (7.5, 1.7) -- (9.1, 1.7);
	\draw[color=gray, dashed] (7.5, 2.4) -- (9.1, 2.4);
	\draw[color=gray, dashed] (7.5, 3.1) -- (9.1, 3.1);
	\draw[color=gray, dashed] (7.5, 3.8) -- (9.1, 3.8);
	\draw[color=gray, dashed] (7.5, 4.5) -- (9.1, 4.5);
	\draw[color=gray, dashed] (7.5, 5.4) -- (9.1, 5.4);
	\draw[color=gray, dashed] (7.5, 6) -- (9.1, 6);
	\draw[color=gray, dashed] (7.5, 6.6) -- (9.1, 6.6);
	\draw[color=gray, dashed] (7.5, 7.2) -- (9.1, 7.2);
\end{tikzpicture}
\end{tabular}
}
\caption{(A)
A realisation of the graph traversed by Hudson's algorithm started from a
sample of three chromosomes with $m$ discrete sites each at time $t = 0$, and
propagated until time $T$. The MRCA on the genetic interval $[v, w)$ is reached
at event \noderef{b}, while that on $[0, v)$ is reached at event \noderef{c}.
The non-ancestral segment $[v, w)$ above
\noderef{a} contributes to the rate of effective recombinations because it
is trapped between ancestral segments. The two columns titled CA and RE
are the respective rates of mergers and recombinations when
the recombination rate is $\rho$.
(B) A corresponding realisation of a Big ARG, which augments Hudson's algorithm
by tracking non-ancestral lineages. The result is a simpler state space and
dynamics, at the cost of extra nodes and edges, highlighted in red, which do
not affect the local tree at any site.
Recombination positions are labelled alphabetically in time, and their ordering
along the genome is $y < v < x < w$, of which the first only appears in panel B.
There are two separate recombination events at link $w$.
}
\label{hudson_vs_bigARG}
\end{figure}

Figure \ref{hudson_vs_bigARG} illustrates the more complex state space
of the Little ARG process, as well as the extra events which occur in the Big ARG process.
Moreover, it depicts the rates of common ancestors and recombination events in each
interval of time of the realisations.
In order to evaluate these rates, e.g.\ for likelihood-based inference
\citep{baumdicker2021efficient, mahmoudi2022bayesian},
it is necessary to know the number of lineages and number of extant links
available for recombination in each time interval.
Some representations may not provide this information.
For example, in the gARG encoding depicted in Figure \ref{fig-ancestry-resolution}C,
it is clear that a recombination takes place between nodes \noderef{i}, \noderef{k} and \noderef{j}.
But the exact time of the recombination event is ambiguous: it could take place at any time between node \noderef{i} and its parents and produce the same gARG.
Because a recombination increases the number of extant lineages by one (in the rootward direction of time), the number of lineages during the same time interval is ambiguous as well.
In fact, this information cannot be recovered from the gARG encoding used in Figure \ref{fig-ancestry-resolution}C without an extrinsic convention.
For the coalescent,
it is sufficient to create \emph{two} gARG nodes at the time of the recombination event, with the interpretation that the two rootward edges from node \noderef{i} in Figure \ref{fig-ancestry-resolution}C belong to the same lineage until the time of nodes \noderef{k} and \noderef{j}, and split into two separate lineages at that time point.
Similarly, the trapped, non-ancestral links along that lineage remain available for effective recombination (i.e.\ one which splits up ancestral material) for the same time interval.
This interpretation is highlighted in
Figure \ref{hudson_vs_bigARG} by drawing only one vertical edge between a recombinant child and its two parents.
